Tidy data structure to support exploration and modeling of temporal data

### 2018-11-27 main.Rmd
roughly done with section on "contextual semantics"

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represents time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" serves a similar purpose as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database.

Table \@ref(tab:tb-sub) shows a "tidy" temporal data example using a subset of tuberculosis cases estimated by @tb-data. It contains 12 observations and 5 variables arranged in a "long" tabular form. Each observation uniquely records the number of people, who are diagnosed tuberculosis for each gender at three selected countries in the years of 2011 and 2012, where column `year` is declared as the index variable, columns `country`, `continent` and `gender` as key. Detailed discussions about index and key specifications will be given in Section \@ref(sec:index) and \@ref(sec:key). Although column `count` is the only measure in this case, it is sufficiently flexible to hold other measured variables, for example, adding the corresponding population size (if known) in order to calibrate the count later.

This high-level data abstraction is semantically structured, which shed lights on time index and what we call "key".

@@

(ref:tb-sub) A small subset of estimates of tuberculosis burden generated by World Health Organisation in 2011 and 2012, with 12 observations and 5 variables. The index refers to column `year`, the key to multiple columns: `country` nested under `continent` crossed with `gender`, and the measured variable to column `count`.

@@

A number of time classes are supported and easily extensible to new types. The base line is that time indices can be potentially ordered in past to future. Characters cannot be ordered, so a variable hosting characters will make it unsupported for index. Ordered factors, however, can be immediately recognized as valid index. For instance, month names have a natural ordering from January to December, and so are weekdays from Monday to Sunday.

@@

The "key" specification contributes to creating a tsibble alongside the index. The concept of "key" is introduced to identify units or subjects that are observed over time in a data table. In the wide format, each column holds a series of values, so that the column itself serves for the sake of identification. In the long format, all column names are melt to the corresponding "key" values. But the "key" is much more flexible than simply column names. Because it is not constrained to a single field, but can be comprised of multiple fields. The identifying variables that the "key" is constituted of, stay the same as they are in the original table, with no further tweaks.

@@

@@

@@

The dataset of 2017 on-time performance for US domestic flights represents event-driven data caught in the wild, sourced from US Bureau of Transportation Statistics (reference). It contains `r format(NROW(flights), big.mark = ",")` operating flights with many measurements (such as departure delay, arrival delay in minutes, and other performance metrics) and detailed flight information (such as origin, destination, plane number and etc.) in a tabular format. This kind of event describes each flight scheduled for departure at a time point in its local time zone. Every single flight would be uniquely identified by the flight number and its scheduled departure time, from a passenger's point of view. In fact, it fails to pass the tsibble hurdle due to duplicates in the original data. An error is immediately raised when attempting to convert this data into a tsibble, and a closer inspection has to be carried out to locate the issue. The **tsibble** package provides the tool to easily find the duplicates in the data with `find_duplicates()`. Below shows the problematic entries.

>>>>

The choice of tidy representation of temporal data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. Contextual semantics are introduced to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series model inputs. Index, key and time interval are the three stone pillars to this new semantically-structured temporal data, which we name as "tsibble".

@@

(ref:tb-sub) A small subset of estimates of tuberculosis burden generated by World Health Organisation in 2011 and 2012, with 12 observations and 5 variables. The index refers to column `year`, the key to multiple columns: `country` and `gender`, and the measured variable to column `count`.

Table \@ref(tab:tb-sub) presents a subset of tuberculosis cases estimated by @tb-data. It contains 12 observations and 5 variables arranged in a "long" tabular form. Each observation hosts the number of people who are diagnosed tuberculosis for each gender at three selected countries in the years of 2011 and 2012. To turn this data into a tsibble, (1) column `year` is declared as the index variable; (2) the key can be made up of columns `country` and `gender`. Despite column `count` is the only measurement in this case, it is sufficiently flexible to hold other measured variables, for example, adding the corresponding population size (if known) in order to calibrate the count later.

@@

A number of time classes are supported and easily extensible to new types. The baseline is that time indices can be potentially ordered from past to future. Characters cannot be ordered, so a variable hosting characters will make it unsupported for index. Ordered factors, however, can be immediately recognized as valid index. For instance, month names have a natural ordering from January to December, and so are weekdays from Monday to Sunday.

@@

The "key" specification contributes to creating a tsibble alongside the index. The concept of "key" is introduced to uniquely identify units or subjects that are observed over time in a data table, which is similar to a primary key [@codd_relational_1970] defining each observation in a relational database. In the wide format, each column holds a series of values, so that the column itself serves for the sake of identification. In the long format, all column names are melt to the corresponding "key" values. But the "key" is much more flexible than simply column names. Because it is not constrained to a single field, but can be comprised of multiple fields. The identifying variables that the "key" is constituted of, stay the same as they are in the original table, with no further tweaks.

@@

## Data first

@@

data first, function starts with the first argument "data", chained with `%>%`. print display.

@@

The dataset of 2017 on-time performance for US domestic flights represents event-driven data caught in the wild, sourced from US Bureau of Transportation Statistics (reference). It contains `r format(NROW(flights), big.mark = ",")` operating flights with many measurements (such as departure delay, arrival delay in minutes, and other performance metrics) and detailed flight information (such as origin, destination, plane number and etc.) in a tabular format. This kind of event describes each flight scheduled for departure at a time point in its local time zone. Every single flight would be uniquely identified by the flight number and its scheduled departure time, from a passenger's point of view. In fact, it fails to pass the tsibble hurdle due to duplicates in the original data. An error is immediately raised when attempting to convert this data into a tsibble, and a closer inspection has to be carried out to locate the issue. The **tsibble** package provides tools to easily locate the duplicates in the data with `duplicates()`. Below shows the problematic entries.
